---
title: "QC & Exploratory Analysis — Orbital Mus"
author: "Orbital Mus Pipeline"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
    theme: flatly
params:
  proj_root: !r rprojroot::find_root(rprojroot::is_git_root)
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, warning = TRUE, message = FALSE,
  fig.width = 10, fig.height = 6
)
```

```{r libraries}
library(DESeq2)
library(ggplot2)
library(pheatmap)
library(jsonlite)
library(dplyr)
```

```{r load-data}
proj_root <- params$proj_root
cfg <- fromJSON(file.path(proj_root, "scripts", "setup", "config.json"))
proc_dir  <- file.path(proj_root, cfg$processed_dir)

heart_counts   <- readRDS(file.path(proc_dir, "heart_counts.rds"))
heart_coldata  <- readRDS(file.path(proc_dir, "heart_coldata.rds"))
organoid_counts  <- readRDS(file.path(proc_dir, "organoid_counts.rds"))
organoid_coldata <- readRDS(file.path(proc_dir, "organoid_coldata.rds"))

min_count <- cfg$qc_min_count
```

This report evaluates QC metrics for two bulk RNA-seq datasets from the Orbital Mus project:

- **Heart tissue**: `r ncol(heart_counts)` mouse samples, 2×2×2 factorial (HU/NL × IR/Sham × KMP/Ctrl)
- **Cardiac organoid**: `r ncol(organoid_counts)` human samples, multi-factorial (Beam × Treatment × Timepoint × Batch)

QC is diagnostic only — no filtered or normalized matrices are saved.

---

# Heart Tissue {.tabset}

## Library Sizes

```{r heart-libsize, fig.height=5}
heart_libsize <- data.frame(
  sample = colnames(heart_counts),
  total  = colSums(heart_counts),
  group  = heart_coldata$group,
  stringsAsFactors = FALSE
)

# IQR outlier threshold
q1  <- quantile(heart_libsize$total, 0.25)
iqr <- IQR(heart_libsize$total)
outlier_threshold <- q1 - 1.5 * iqr
heart_libsize$outlier <- heart_libsize$total < outlier_threshold

n_outliers_heart_lib <- sum(heart_libsize$outlier)

ggplot(heart_libsize, aes(x = reorder(sample, total), y = total / 1e6, fill = group)) +
  geom_col() +
  geom_hline(yintercept = outlier_threshold / 1e6, linetype = "dashed", color = "red") +
  coord_flip() +
  labs(
    title = "Heart: Per-Sample Library Size",
    x = NULL, y = "Total counts (millions)",
    caption = paste0("Red line: IQR outlier threshold (Q1 - 1.5×IQR = ",
                     format(round(outlier_threshold), big.mark = ","), ")")
  ) +
  theme_minimal(base_size = 10) +
  theme(axis.text.y = element_text(size = 5))
```

`r if (n_outliers_heart_lib > 0) paste0("**⚠ ", n_outliers_heart_lib, " sample(s) flagged as library-size outliers:** ", paste(heart_libsize$sample[heart_libsize$outlier], collapse = ", ")) else "No library-size outliers detected."`

## Gene Filtering

```{r heart-filter}
# Keep genes with >= min_count in at least N samples (N = smallest group size)
heart_min_samples <- min(table(heart_coldata$group))
heart_keep <- rowSums(heart_counts >= min_count) >= heart_min_samples
heart_n_keep <- sum(heart_keep)
heart_n_remove <- sum(!heart_keep)

cat(sprintf(
  "Gene filtering (>=%d counts in >=%d samples):\n  Retained: %s genes\n  Removed:  %s genes (%.1f%%)\n",
  min_count, heart_min_samples,
  format(heart_n_keep, big.mark = ","),
  format(heart_n_remove, big.mark = ","),
  100 * heart_n_remove / nrow(heart_counts)
))
```

## VST + PCA {.tabset}

```{r heart-vst}
heart_counts_filt <- heart_counts[heart_keep, ]
heart_dds <- DESeqDataSetFromMatrix(
  countData = heart_counts_filt,
  colData   = heart_coldata,
  design    = ~ 1
)
heart_vst <- vst(heart_dds, blind = TRUE)
```

### PCA — HU

```{r heart-pca-hu, fig.height=5}
plotPCA(heart_vst, intgroup = "HU") +
  ggtitle("Heart PCA: HU (Hindlimb Unloading)") +
  theme_minimal()
```

### PCA — IR

```{r heart-pca-ir, fig.height=5}
plotPCA(heart_vst, intgroup = "IR") +
  ggtitle("Heart PCA: IR (Irradiation)") +
  theme_minimal()
```

### PCA — KMP

```{r heart-pca-kmp, fig.height=5}
plotPCA(heart_vst, intgroup = "KMP") +
  ggtitle("Heart PCA: KMP (Treatment)") +
  theme_minimal()
```

### PCA — Group

```{r heart-pca-group, fig.height=5}
plotPCA(heart_vst, intgroup = "group") +
  ggtitle("Heart PCA: Full Group") +
  theme_minimal()
```

## Sample Distance Heatmap

```{r heart-heatmap, fig.height=8, fig.width=9}
heart_dist <- dist(t(assay(heart_vst)))
heart_dist_mat <- as.matrix(heart_dist)

heart_anno <- data.frame(
  HU  = heart_coldata$HU,
  IR  = heart_coldata$IR,
  KMP = heart_coldata$KMP,
  row.names = rownames(heart_coldata)
)

pheatmap(
  heart_dist_mat,
  clustering_distance_rows = heart_dist,
  clustering_distance_cols = heart_dist,
  annotation_col = heart_anno,
  main = "Heart: Sample-to-Sample Euclidean Distance",
  fontsize_row = 5, fontsize_col = 5
)
```

## Outlier Summary

```{r heart-outliers}
# PCA-based outlier detection: flag samples > 3 SD from centroid on PC1/PC2
heart_pca_data <- plotPCA(heart_vst, intgroup = "group", returnData = TRUE)
heart_centroid <- c(mean(heart_pca_data$PC1), mean(heart_pca_data$PC2))
heart_pca_data$dist_to_centroid <- sqrt(
  (heart_pca_data$PC1 - heart_centroid[1])^2 +
  (heart_pca_data$PC2 - heart_centroid[2])^2
)
pca_mean <- mean(heart_pca_data$dist_to_centroid)
pca_sd   <- sd(heart_pca_data$dist_to_centroid)
heart_pca_data$pca_outlier <- heart_pca_data$dist_to_centroid > (pca_mean + 3 * pca_sd)

heart_outlier_summary <- data.frame(
  sample = heart_libsize$sample,
  library_size_outlier = heart_libsize$outlier,
  pca_outlier = heart_pca_data$pca_outlier[match(heart_libsize$sample, heart_pca_data$name)]
)
heart_outlier_summary$any_flag <- heart_outlier_summary$library_size_outlier |
                                  heart_outlier_summary$pca_outlier

flagged <- heart_outlier_summary[heart_outlier_summary$any_flag, ]
if (nrow(flagged) > 0) {
  cat("Flagged heart samples:\n")
  print(flagged)
} else {
  cat("No heart samples flagged as outliers.\n")
}
```

---

# Cardiac Organoid {.tabset}

## Library Sizes

```{r organoid-libsize, fig.height=5}
organoid_libsize <- data.frame(
  sample = colnames(organoid_counts),
  total  = colSums(organoid_counts),
  group  = organoid_coldata$group,
  batch  = organoid_coldata$batch,
  stringsAsFactors = FALSE
)

q1_o  <- quantile(organoid_libsize$total, 0.25)
iqr_o <- IQR(organoid_libsize$total)
outlier_threshold_o <- q1_o - 1.5 * iqr_o
organoid_libsize$outlier <- organoid_libsize$total < outlier_threshold_o

n_outliers_org_lib <- sum(organoid_libsize$outlier)

ggplot(organoid_libsize, aes(x = reorder(sample, total), y = total / 1e6, fill = group)) +
  geom_col() +
  geom_hline(yintercept = outlier_threshold_o / 1e6, linetype = "dashed", color = "red") +
  coord_flip() +
  labs(
    title = "Organoid: Per-Sample Library Size",
    x = NULL, y = "Total counts (millions)",
    caption = paste0("Red line: IQR outlier threshold (Q1 - 1.5×IQR = ",
                     format(round(outlier_threshold_o), big.mark = ","), ")")
  ) +
  theme_minimal(base_size = 10) +
  theme(axis.text.y = element_text(size = 6))
```

`r if (n_outliers_org_lib > 0) paste0("**⚠ ", n_outliers_org_lib, " sample(s) flagged as library-size outliers:** ", paste(organoid_libsize$sample[organoid_libsize$outlier], collapse = ", ")) else "No library-size outliers detected."`

## Gene Filtering

```{r organoid-filter}
organoid_min_samples <- min(table(organoid_coldata$group))
organoid_keep <- rowSums(organoid_counts >= min_count) >= organoid_min_samples
organoid_n_keep <- sum(organoid_keep)
organoid_n_remove <- sum(!organoid_keep)

cat(sprintf(
  "Gene filtering (>=%d counts in >=%d samples):\n  Retained: %s genes\n  Removed:  %s genes (%.1f%%)\n",
  min_count, organoid_min_samples,
  format(organoid_n_keep, big.mark = ","),
  format(organoid_n_remove, big.mark = ","),
  100 * organoid_n_remove / nrow(organoid_counts)
))
```

## VST + PCA {.tabset}

```{r organoid-vst}
organoid_counts_filt <- organoid_counts[organoid_keep, ]
organoid_dds <- DESeqDataSetFromMatrix(
  countData = organoid_counts_filt,
  colData   = organoid_coldata,
  design    = ~ 1
)
organoid_vst <- vst(organoid_dds, blind = TRUE)
```

### PCA — Batch

```{r organoid-pca-batch, fig.height=5}
plotPCA(organoid_vst, intgroup = "batch") +
  ggtitle("Organoid PCA: Batch") +
  theme_minimal()
```

### PCA — Beam

```{r organoid-pca-beam, fig.height=5}
plotPCA(organoid_vst, intgroup = "beam") +
  ggtitle("Organoid PCA: Beam (Irradiation)") +
  theme_minimal()
```

### PCA — Treatment

```{r organoid-pca-treatment, fig.height=5}
plotPCA(organoid_vst, intgroup = "treatment") +
  ggtitle("Organoid PCA: Treatment (KMP/Control)") +
  theme_minimal()
```

### PCA — Timepoint

```{r organoid-pca-timepoint, fig.height=5}
plotPCA(organoid_vst, intgroup = "timepoint") +
  ggtitle("Organoid PCA: Timepoint (Day2/Day9)") +
  theme_minimal()
```

## Sample Distance Heatmap

```{r organoid-heatmap, fig.height=8, fig.width=9}
organoid_dist <- dist(t(assay(organoid_vst)))
organoid_dist_mat <- as.matrix(organoid_dist)

organoid_anno <- data.frame(
  Batch     = organoid_coldata$batch,
  Beam      = organoid_coldata$beam,
  Treatment = organoid_coldata$treatment,
  Timepoint = organoid_coldata$timepoint,
  row.names = rownames(organoid_coldata)
)

pheatmap(
  organoid_dist_mat,
  clustering_distance_rows = organoid_dist,
  clustering_distance_cols = organoid_dist,
  annotation_col = organoid_anno,
  main = "Organoid: Sample-to-Sample Euclidean Distance",
  fontsize_row = 6, fontsize_col = 6
)
```

## Outlier Summary

```{r organoid-outliers}
organoid_pca_data <- plotPCA(organoid_vst, intgroup = "group", returnData = TRUE)
org_centroid <- c(mean(organoid_pca_data$PC1), mean(organoid_pca_data$PC2))
organoid_pca_data$dist_to_centroid <- sqrt(
  (organoid_pca_data$PC1 - org_centroid[1])^2 +
  (organoid_pca_data$PC2 - org_centroid[2])^2
)
org_pca_mean <- mean(organoid_pca_data$dist_to_centroid)
org_pca_sd   <- sd(organoid_pca_data$dist_to_centroid)
organoid_pca_data$pca_outlier <- organoid_pca_data$dist_to_centroid > (org_pca_mean + 3 * org_pca_sd)

organoid_outlier_summary <- data.frame(
  sample = organoid_libsize$sample,
  library_size_outlier = organoid_libsize$outlier,
  pca_outlier = organoid_pca_data$pca_outlier[match(organoid_libsize$sample, organoid_pca_data$name)]
)
organoid_outlier_summary$any_flag <- organoid_outlier_summary$library_size_outlier |
                                     organoid_outlier_summary$pca_outlier

flagged_org <- organoid_outlier_summary[organoid_outlier_summary$any_flag, ]
if (nrow(flagged_org) > 0) {
  cat("Flagged organoid samples:\n")
  print(flagged_org)
} else {
  cat("No organoid samples flagged as outliers.\n")
}
```

---

# Session Info

```{r session-info}
sessionInfo()
```
