---
title: "Deconvolution, Differential Expression, and GSEA Summary"
author: "Orbital Mus Pipeline"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
    theme: flatly
params:
  proj_root: "."
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, warning = TRUE, message = FALSE,
  fig.width = 10, fig.height = 6
)
```

```{r libraries}
library(DESeq2)
library(ggplot2)
library(dplyr)
library(tidyr)
library(stringr)
library(RColorBrewer)
library(jsonlite)
library(ggrepel)
library(ggrastr)
library(patchwork)
library(ComplexUpset)
library(kableExtra)
library(gt)
```

```{r load-data}
proj_root <- params$proj_root
cfg <- fromJSON(file.path(proj_root, "scripts", "setup", "config.json"))
proc_dir <- file.path(proj_root, cfg$processed_dir)

refs <- cfg$decon_references
alpha <- cfg$de_alpha

# B6J-matched references for DE and GSEA
b6j_refs <- c("bulk_decon", "mortazavi_b6j")

compositions <- lapply(refs, function(ref) {
  read.csv(file.path(proc_dir, "decon", paste0(ref, "_compositions.csv"))) |>
    mutate(reference = ref)
}) |> bind_rows()

comparisons <- lapply(refs, function(ref) {
  read.csv(file.path(proc_dir, "models", paste0(ref, "_de_comparison.csv"))) |>
    mutate(reference = ref)
}) |> bind_rows()

markers <- lapply(refs, function(ref) {
  read.csv(file.path(proc_dir, "decon", paste0(ref, "_markers.csv"))) |>
    mutate(reference = ref)
}) |> bind_rows()

clr_covariates <- lapply(refs, function(ref) {
  f <- file.path(proc_dir, "models", paste0(ref, "_clr_covariates.csv"))
  if (file.exists(f)) read.csv(f) else NULL
}) |> bind_rows()

dirichlet_coefs <- lapply(refs, function(ref) {
  f <- file.path(proc_dir, "models", paste0(ref, "_dirichlet_coefficients.csv"))
  if (file.exists(f)) read.csv(f) |> mutate(reference = ref) else NULL
}) |> bind_rows()

contrast_indices <- lapply(refs, function(ref) {
  f <- file.path(proc_dir, "models", paste0(ref, "_contrast_index.csv"))
  if (file.exists(f)) read.csv(f) |> mutate(reference = ref) else NULL
}) |> bind_rows()

# Load DE results RDS for volcanos and UpSet
results_adj <- lapply(refs, function(ref) {
  f <- file.path(proc_dir, "models", paste0(ref, "_results_adjusted.rds"))
  if (file.exists(f)) readRDS(f) else NULL
})
names(results_adj) <- refs

results_unadj <- lapply(refs, function(ref) {
  f <- file.path(proc_dir, "models", paste0(ref, "_results_unadjusted.rds"))
  if (file.exists(f)) readRDS(f) else NULL
})
names(results_unadj) <- refs

# Gene name map
gene_map <- read.csv(file.path(proc_dir, "heart_gene_names.csv"))

# Load GSEA summaries
gsea_summaries <- lapply(refs, function(ref) {
  lapply(c("adjusted", "unadjusted"), function(model) {
    f <- file.path(proc_dir, "gsea", paste0(ref, "_", model, "_combined.csv"))
    if (file.exists(f)) read.csv(f) else NULL
  }) |> bind_rows()
}) |> bind_rows()

# Load DDS objects for CLR covariate significance
dds_adj_list <- lapply(b6j_refs, function(ref) {
  f <- file.path(proc_dir, "models", paste0(ref, "_dds_adjusted.rds"))
  if (file.exists(f)) readRDS(f) else NULL
})
names(dds_adj_list) <- b6j_refs

# ---- Global rename: bulk_decon -> Rau2025 ----
rename_ref <- function(x) gsub("bulk_decon", "Rau2025", x)

compositions$reference     <- rename_ref(compositions$reference)
comparisons$reference      <- rename_ref(comparisons$reference)
markers$reference          <- rename_ref(markers$reference)
clr_covariates$reference   <- rename_ref(clr_covariates$reference)
dirichlet_coefs$reference  <- rename_ref(dirichlet_coefs$reference)
contrast_indices$reference <- rename_ref(contrast_indices$reference)
gsea_summaries$reference   <- rename_ref(gsea_summaries$reference)
refs <- rename_ref(refs)
b6j_refs <- rename_ref(b6j_refs)
names(results_adj)   <- rename_ref(names(results_adj))
names(results_unadj) <- rename_ref(names(results_unadj))
names(dds_adj_list)  <- rename_ref(names(dds_adj_list))

# ---- Extract per-gene CLR covariate significance ----
clr_sig_list <- list()
for (ref in b6j_refs) {
  dds <- dds_adj_list[[ref]]
  if (is.null(dds)) next

  clr_ref <- clr_covariates |> filter(reference == ref)
  clr_names <- clr_ref$clr_name
  ct_names <- clr_ref$celltype

  rn <- resultsNames(dds)
  clr_coefs <- intersect(clr_names, rn)
  if (length(clr_coefs) != 2) next

  res1 <- as.data.frame(results(dds, name = clr_coefs[1]))
  res2 <- as.data.frame(results(dds, name = clr_coefs[2]))

  ct1_label <- paste0(ct_names[clr_names == clr_coefs[1]], " only")
  ct2_label <- paste0(ct_names[clr_names == clr_coefs[2]], " only")

  clr_sig_list[[ref]] <- data.frame(
    gene_id = rownames(res1),
    reference = ref,
    sig_ct1 = !is.na(res1$padj) & res1$padj <= alpha,
    sig_ct2 = !is.na(res2$padj) & res2$padj <= alpha
  ) |> mutate(
    ct_assoc = case_when(
      sig_ct1 & sig_ct2 ~ "Both cell types",
      sig_ct1 ~ ct1_label,
      sig_ct2 ~ ct2_label,
      TRUE ~ "Neither"
    )
  ) |> select(gene_id, reference, ct_assoc)
}

clr_sig_all <- bind_rows(clr_sig_list)
comparisons <- comparisons |>
  left_join(clr_sig_all, by = c("gene_id", "reference"))
```

```{r palettes}
# Group palette
group_pal <- brewer.pal(8, "Dark2")
names(group_pal) <- cfg$heart_group_order

# Key contrasts for focused plots
key_contrasts <- c(
  "HU_vs_NL_MAIN_EFFECT", "IR_vs_Sham_MAIN_EFFECT", "KMP_vs_Sham_MAIN_EFFECT",
  "Interaction_KMP_x_HU", "Interaction_KMP_x_IR", "Interaction_HU_x_IR"
)

# Readable contrast labels (expanded to include pairwise)
contrast_labels <- c(
  "HU_vs_NL_MAIN_EFFECT"    = "HU vs NL",
  "IR_vs_Sham_MAIN_EFFECT"  = "IR vs Sham",
  "KMP_vs_Sham_MAIN_EFFECT" = "KMP vs Ctrl",
  "Interaction_KMP_x_HU"    = "KMP x HU",
  "Interaction_KMP_x_IR"    = "KMP x IR",
  "Interaction_HU_x_IR"     = "HU x IR",
  "HU_vs_Ctrl"              = "HU vs Ctrl (pairwise)",
  "IR_vs_Ctrl"              = "IR vs Ctrl (pairwise)",
  "KMP_vs_Ctrl"             = "KMP vs Ctrl (pairwise)"
)

# Consistent cell type colors across references via family mapping
celltype_families <- list(
  cardiomyocyte = c("Cardiomyocytes", "muscle",
                    "ventricular cardiac myocyte",
                    "atrial cardiac myocyte"),
  fibroblast = c("Fibroblast", "stromal", "fibroblast"),
  immune = c("Macrophage", "immune", "macrophage", "B cell", "T cell"),
  pericyte = c("Pericytes/SMC", "pericyte", "smooth muscle cell"),
  endothelial = c("endothelial", "endothelial cell",
                  "lymphatic endothelial cell", "endocardial cell"),
  neural = c("neural", "neuron", "Schwann cell"),
  adipocyte = c("adipocyte")
)

family_colors <- c(
  cardiomyocyte = "#66C2A5", fibroblast = "#FC8D62", immune = "#8DA0CB",
  pericyte = "#E78AC3", endothelial = "#A6D854", neural = "#FFD92F",
  adipocyte = "#E5C494"
)

# Map each cell type to its family color
all_celltypes <- unique(compositions$CellType)
celltype_pal <- setNames(rep("#B3B3B3", length(all_celltypes)), all_celltypes)
for (fam in names(celltype_families)) {
  for (ct in all_celltypes) {
    if (any(sapply(celltype_families[[fam]], function(pat) grepl(pat, ct, ignore.case = TRUE)))) {
      celltype_pal[ct] <- family_colors[fam]
    }
  }
}

# Assign each cell type to a family name for the reference comparison
assign_family <- function(ct) {
  for (fam in names(celltype_families)) {
    if (any(sapply(celltype_families[[fam]], function(pat) grepl(pat, ct, ignore.case = TRUE)))) {
      return(fam)
    }
  }
  return("other")
}

# Base theme for all plots
theme_report <- function(base_size = 10) {
  theme_minimal(base_size = base_size) +
    theme(
      text = element_text(size = 9),
      panel.grid.minor = element_blank(),
      plot.background = element_rect(fill = "transparent", color = NA),
      legend.background = element_rect(fill = "transparent")
    )
}

# Helper for styled kable tables
styled_kable <- function(df, ...) {
  knitr::kable(df, ...) |>
    kable_styling(
      bootstrap_options = c("striped", "hover", "condensed"),
      full_width = FALSE, font_size = 12
    )
}
```

# Introduction

## Biological Question

Hindlimb unloading (HU), ionizing radiation (IR), and the countermeasure KMP
model distinct spaceflight stressors on the murine heart. The core question:
**which expression changes are intrinsic transcriptional responses vs. secondary
consequences of cell type composition shifts?**

We use two independent scRNA-seq reference datasets for deconvolution:

- **Rau2025** ([Gural et al. 2025](https://doi.org/10.1371/journal.pgen.1011807)):
  4 major cell types from C57BL/6J mouse hearts
- **Mortazavi** (IGVF Mortazavi lab): multi-strain mouse heart atlas with three
  granularity levels -- **general** (8 types), **celltype** (14 types), and
  **b6j** (8 B6J-specific types)

We select Mortazavi general as the primary non-Rau reference because it offers
richer diversity in expression profiles (8 cell types vs. 4 in Rau2025) without
the over-splitting of 14 types in the celltype version, which risks poorly
resolved signatures.

## Experimental Design

```{r design-table}
design_df <- data.frame(
  Factor = c("Loading", "Radiation", "Countermeasure"),
  Levels = c("HU / NL", "IR / Sham", "KMP / Ctrl"),
  Rationale = c(
    "Simulates microgravity-induced cardiac unloading",
    "Mimics space radiation exposure",
    "Tests a pharmacological countermeasure"
  )
)
styled_kable(design_df, caption = "2x2x2 factorial design (~80 samples, 8 groups, ~10/group)")
```

## Analytical Strategy

1. **MuSiC deconvolution** -- estimate cell type proportions from bulk RNA-seq
   using several single-cell/nucleus reference panels (15 markers per cell type),
   then select one primary reference for downstream analysis
2. **Dirichlet regression** -- test whether treatments shift cell type
   composition (`CellTypes ~ HU * IR * KMP`)
3. **CLR-adjusted DE** -- include CLR-transformed proportions of the top 2 most
   abundant cell types as additive covariates in DESeq2 (full interaction model
   for treatments: `HU * IR * KMP`). Compare how gene significance changes
   with and without cell type covariates
4. **GSEA** -- identify enriched pathways in adjusted ranked gene lists for each
   treatment contrast and interaction, focusing on the most-changed genes

Source: [github.com/guralbrian/orbital_mus](https://github.com/guralbrian/orbital_mus)

---

# Methods Overview

## CLR Covariate Selection

The composition-adjusted DESeq2 model includes CLR-transformed proportions
of the **top 2 most abundant cell types** (by mean proportion across all samples)
as covariates. This limits the number of added covariates to avoid overfitting,
particularly for references with many cell types.

```{r clr-table}
if (nrow(clr_covariates) > 0) {
  styled_kable(
    clr_covariates |> select(reference, celltype),
    caption = "Top-2 CLR covariates per reference"
  )
}
```

## Model Formulas

- **Unadjusted DE:** `~ 0 + HU * IR * KMP`
- **Adjusted DE:** `~ 0 + group + clr.CellType1 + clr.CellType2`
- **Dirichlet regression:** `CellTypes ~ HU * IR * KMP`

Both DE models use cell-means parameterization with 19 numeric contrast vectors
(10 pairwise, 3 main effects, 3 marginal means, 3 two-way interactions).
LFC shrinkage via ashr.

## GSEA Ranking Metric

Genes are ranked by **signed -log10(p-value)**: `-log10(pvalue) * sign(log2FoldChange)`.
This weights both statistical significance and effect direction, matching the
approach in [Gural et al. 2025](https://doi.org/10.1371/journal.pgen.1011807).

GSEA was run using clusterProfiler's `gseGO()` (GO Biological Process, ENSEMBL IDs)
and `gseKEGG()` (KEGG pathways, ENTREZID). Parameters: minGSSize=10, maxGSSize=500,
BH adjustment, eps=0.

---

# Reference Panel

```{r ref-panel-table}
ref_sources <- data.frame(
  reference = c("Rau2025", "mortazavi_general", "mortazavi_celltype", "mortazavi_b6j"),
  Source = c(
    "Gural et al. 2025",
    "IGVF Mortazavi lab",
    "IGVF Mortazavi lab",
    "IGVF Mortazavi lab"
  ),
  Strain = c("C57BL/6J", "Multi-strain", "Multi-strain", "C57BL/6J"),
  Granularity = c("4 major types", "8 general types", "14 granular types", "8 B6J types")
)

marker_summary <- markers |>
  group_by(reference) |>
  summarise(
    n_celltypes = n_distinct(celltype),
    n_markers = n(),
    .groups = "drop"
  )

ref_table <- ref_sources |>
  left_join(marker_summary, by = "reference") |>
  mutate(
    Role = ifelse(reference %in% b6j_refs,
                  "Composition + DE + GSEA",
                  "Composition + Dirichlet only")
  ) |>
  dplyr::rename(Reference = reference, `N Cell Types` = n_celltypes, `N Markers` = n_markers)

styled_kable(ref_table, caption = "Single-cell/nucleus reference panels")
```

---

# What Are the Cell Type Compositions? {.tabset}

## Stacked Bars {.tabset .tabset-pills}

```{r comp-stacked, results='asis', fig.width=14, fig.height=9}
for (ref in refs) {
  cat("\n\n###", ref, "\n\n")

  comp_ref <- compositions |> filter(reference == ref)

  cell_order <- comp_ref |>
    group_by(CellType) |>
    summarise(mean_prop = mean(Prop, na.rm = TRUE), .groups = "drop") |>
    arrange(desc(mean_prop)) |>
    pull(CellType)

  ct_pal <- celltype_pal[cell_order]

  sample_order <- comp_ref |>
    group_by(group, sample_id) |>
    slice_max(Prop, n = 1, with_ties = FALSE) |>
    ungroup() |>
    arrange(group, CellType, desc(Prop)) |>
    pull(sample_id)

  comp_ref <- comp_ref |>
    mutate(
      sample_id = factor(sample_id, levels = sample_order),
      group = factor(group, levels = cfg$heart_group_order),
      CellType = factor(CellType, levels = cell_order)
    )

  p_bar <- ggplot(comp_ref, aes(x = sample_id, y = Prop, fill = CellType)) +
    geom_col(color = "black", linewidth = 0.1) +
    facet_wrap(~ group, ncol = 4, scales = "free_x") +
    scale_fill_manual(values = ct_pal,
                      labels = function(x) str_wrap(x, width = 15)) +
    labs(x = NULL, y = "Proportion", fill = "Cell Type") +
    theme_report() +
    theme(
      axis.text.x = element_text(angle = 90, hjust = 1, size = 6),
      strip.text = element_text(size = 9)
    )
  print(p_bar)
  cat("\n\n")
}
```

## Boxplots by Group {.tabset .tabset-pills}

```{r comp-boxplots, results='asis', fig.width=12, fig.height=8}
for (ref in refs) {
  cat("\n\n###", ref, "\n\n")

  comp_ref <- compositions |>
    filter(reference == ref) |>
    mutate(
      group = factor(group, levels = cfg$heart_group_order),
      Pct = Prop * 100
    )

  p_box <- ggplot(comp_ref, aes(x = group, y = Pct, fill = group)) +
    geom_boxplot(outlier.size = 0.8, linewidth = 0.3) +
    geom_jitter(width = 0.15, alpha = 0.4, size = 0.8) +
    facet_wrap(~ CellType, scales = "free_y") +
    scale_fill_manual(values = group_pal) +
    labs(x = NULL, y = "Proportion (%)") +
    theme_report() +
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1, size = 7),
      legend.position = "bottom"
    ) +
    guides(fill = guide_legend(nrow = 2))
  print(p_box)
  cat("\n\n")

  # Mean proportions table (percent)
  mean_tbl <- compositions |>
    filter(reference == ref) |>
    mutate(Pct = Prop * 100) |>
    group_by(group, CellType) |>
    summarise(mean_pct = mean(Pct, na.rm = TRUE), .groups = "drop") |>
    pivot_wider(names_from = CellType, values_from = mean_pct)
  print(styled_kable(mean_tbl, digits = 1,
                     caption = paste0("Mean proportions (%) -- ", ref)))
  cat("\n\n")
}
```

## Reference Comparison

```{r ref-comparison, fig.width=8, fig.height=5}
# Map cell types to families and compare across references
comp_family <- compositions |>
  mutate(
    family = sapply(CellType, assign_family),
    Pct = Prop * 100
  ) |>
  group_by(reference, sample_id, family) |>
  summarise(Pct = sum(Pct, na.rm = TRUE), .groups = "drop")

ggplot(comp_family, aes(x = family, y = Pct, fill = reference)) +
  geom_boxplot(outlier.size = 0.5, linewidth = 0.3,
               position = position_dodge(width = 0.8)) +
  scale_fill_brewer(palette = "Dark2") +
  labs(x = NULL, y = "Proportion (%)", fill = "Reference") +
  theme_report() +
  theme(axis.text.x = element_text(angle = 30, hjust = 1, size = 9))
```

```{r comp-dynamic-summary, results='asis'}
# Dynamic summary sentence
cm_range <- compositions |>
  mutate(family = sapply(CellType, assign_family)) |>
  filter(family == "cardiomyocyte") |>
  group_by(reference, sample_id) |>
  summarise(cm_prop = sum(Prop, na.rm = TRUE), .groups = "drop") |>
  group_by(reference) |>
  summarise(mean_cm = mean(cm_prop), .groups = "drop")

cm_lo <- round(min(cm_range$mean_cm) * 100, 0)
cm_hi <- round(max(cm_range$mean_cm) * 100, 0)

cat(paste0(
  "Across all references, cardiomyocyte-family cells dominate (",
  cm_lo, "--", cm_hi, "%), consistent with left ventricular tissue.\n\n"
))
```

---

# Do Treatments Shift Composition?

```{r dirichlet-section, results='asis', fig.width=10, fig.height=8}
if (nrow(dirichlet_coefs) > 0) {
  dir_ref <- dirichlet_coefs |> filter(reference == "mortazavi_general")

  if (nrow(dir_ref) > 0) {
    dir_ref <- dir_ref |>
      mutate(Feature_wrap = case_when(
        Feature == "(Intercept)"         ~ "Intercept",
        Feature == "HUHU"                ~ "HU",
        Feature == "IRIR"                ~ "IR",
        Feature == "KMPKMP"              ~ "KMP",
        Feature == "HUHU:IRIR"           ~ "HU x IR",
        Feature == "HUHU:KMPKMP"         ~ "HU x KMP",
        Feature == "IRIR:KMPKMP"         ~ "IR x KMP",
        Feature == "HUHU:IRIR:KMPKMP"    ~ "HU x IR x KMP",
        TRUE                             ~ Feature
      ))

    dir_plot <- dir_ref |> filter(Feature != "(Intercept)")

    p_dir <- ggplot(dir_plot, aes(y = Feature_wrap, x = Estimate,
                                   color = Pr...z..)) +
      geom_errorbarh(
        aes(xmin = Estimate - StdError, xmax = Estimate + StdError),
        height = 0.3, linewidth = 0.5
      ) +
      geom_point(size = 1.6, color = "#474747") +
      geom_vline(xintercept = 0, linetype = "solid", linewidth = 0.5) +
      binned_scale(
        aesthetics = "color", scale_name = "stepsn",
        palette = function(x) c("#FDE725FF", "#73D055FF", "#20A387FF", "#482677FF"),
        breaks = c(0.005, 0.01, 0.05),
        limits = c(0.0005, 0.5),
        show.limits = TRUE, guide = "colorsteps"
      ) +
      facet_wrap(~ CellType, ncol = 3, scales = "free_x") +
      labs(x = "Coefficient estimate", y = NULL, color = "p-value") +
      theme_report() +
      theme(
        strip.text = element_text(size = 8),
        panel.grid.major = element_line(colour = "grey90", linewidth = 0.3)
      )
    print(p_dir)
    cat("\n\n")

    # Significance summary table
    sig_dir <- dir_plot |>
      filter(Pr...z.. < 0.05) |>
      select(CellType, Feature_wrap, Estimate, StdError, Pr...z..) |>
      arrange(Pr...z..)
    if (nrow(sig_dir) > 0) {
      print(styled_kable(sig_dir, digits = 4,
                         caption = "Significant coefficients (p < 0.05) -- mortazavi_general"))
    } else {
      cat("No coefficients with p < 0.05.\n\n")
    }
  } else {
    cat("No Dirichlet results available for mortazavi_general.\n\n")
  }
} else {
  cat("No Dirichlet regression results found.\n\n")
}
```

---

# How Does Composition Adjustment Affect DE? {.tabset .tabset-pills}

We've previously shown that compositional changes in bulk tissue can inflate
false-positive DEGs. To address this, we've included CLR-transformed cell type
proportions as covariates in DESeq2. Points above the diagonal were more
significant after adjustment (signal was masked by composition); points below
were less significant (signal was inflated by composition). Dots are colored
by whether the gene's expression is significantly associated with the CLR
covariate cell types.

**Shown for strain-matched (B6J) references only.**

```{r de-adjustment, results='asis', fig.width=10, fig.height=6}
for (ref in b6j_refs) {
  cat("\n\n##", ref, "\n\n")

  comp_ref <- comparisons |> filter(reference == ref)

  # Get CLR covariate names for this reference
  clr_ref <- clr_covariates |> filter(reference == ref)

  # Scatter: padj adjusted vs unadjusted (key contrasts)
  comp_key <- comp_ref |>
    filter(contrast %in% key_contrasts) |>
    mutate(
      neg_log10_adj = -log10(pmax(padj.adj, 1e-300)),
      neg_log10_unadj = -log10(pmax(padj.unadj, 1e-300)),
      contrast_label = factor(contrast, levels = key_contrasts,
                              labels = contrast_labels[key_contrasts]),
      ct_assoc = ifelse(is.na(ct_assoc), "Neither", ct_assoc)
    )

  if (nrow(comp_key) > 0 && nrow(clr_ref) == 2) {
    # Build dynamic color palette for cell type association
    ct1_label <- paste0(clr_ref$celltype[1], " only")
    ct2_label <- paste0(clr_ref$celltype[2], " only")
    assoc_levels <- c("Neither", ct1_label, ct2_label, "Both cell types")
    assoc_colors <- c("Neither" = "#999999",
                      setNames("#e41a1c", ct1_label),
                      setNames("#377eb8", ct2_label),
                      "Both cell types" = "#8856a7")
    comp_key$ct_assoc <- factor(comp_key$ct_assoc, levels = assoc_levels)

    # Dummy points to force equal axes per panel
    axis_range <- comp_key |>
      group_by(contrast_label) |>
      summarise(
        max_val = max(c(neg_log10_adj, neg_log10_unadj), na.rm = TRUE),
        .groups = "drop"
      )
    dummy <- axis_range |>
      mutate(neg_log10_unadj = max_val, neg_log10_adj = max_val)

    p_dot <- ggplot(comp_key, aes(x = neg_log10_unadj, y = neg_log10_adj,
                                   color = ct_assoc)) +
      geom_blank(data = dummy, aes(x = neg_log10_unadj, y = neg_log10_adj),
                 inherit.aes = FALSE) +
      geom_point_rast(size = 0.3, alpha = 0.3, dpi = 600) +
      geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red") +
      facet_wrap(~ contrast_label, scales = "free") +
      scale_color_manual(values = assoc_colors) +
      labs(x = "-log10(padj) Unadjusted",
           y = "-log10(padj) Adjusted",
           color = "Cell type\nassociation") +
      theme_report()
    print(p_dot)
    cat("\n\n")

    # DEG count table (key contrasts) with renamed columns
    adj_summary <- comp_key |>
      group_by(contrast_label) |>
      summarise(
        `Sig. (adjusted)` = sum(padj.adj <= alpha, na.rm = TRUE),
        `Sig. (unadjusted)` = sum(padj.unadj <= alpha, na.rm = TRUE),
        `Adjusted only` = sum(adj_only_sig, na.rm = TRUE),
        `Unadjusted only` = sum(unadj_only_sig, na.rm = TRUE),
        Both = sum(both_sig, na.rm = TRUE),
        .groups = "drop"
      )
    print(styled_kable(adj_summary, caption = paste0("DEG counts (key contrasts) -- ", ref)))
    cat("\n\n")

    # Impact bar chart: cross-tabulate sig status × cell type association
    impact_ct <- comp_key |>
      mutate(
        sig_status = case_when(
          adj_only_sig  ~ "Adjusted only",
          unadj_only_sig ~ "Unadjusted only",
          both_sig ~ "Both",
          TRUE ~ "Neither"
        )
      ) |>
      filter(sig_status != "Neither") |>
      count(contrast_label, sig_status, ct_assoc) |>
      mutate(sig_status = factor(sig_status,
                                 levels = c("Adjusted only", "Unadjusted only", "Both")))

    if (nrow(impact_ct) > 0) {
      p_impact <- ggplot(impact_ct, aes(x = contrast_label, y = n, fill = ct_assoc)) +
        geom_col(position = "dodge") +
        facet_wrap(~ sig_status) +
        scale_fill_manual(values = assoc_colors) +
        labs(x = NULL, y = "Number of genes", fill = "Cell type\nassociation") +
        theme_report() +
        theme(axis.text.x = element_text(angle = 45, hjust = 1))
      print(p_impact)
      cat("\n\n")
    }
  }
}

# Dynamic summary
adj_impact <- comparisons |>
  filter(reference %in% b6j_refs, contrast %in% key_contrasts) |>
  summarise(
    removed = sum(unadj_only_sig, na.rm = TRUE),
    revealed = sum(adj_only_sig, na.rm = TRUE)
  )
cat(paste0(
  "\n\nComposition adjustment removed ", adj_impact$removed,
  " DEGs and revealed ", adj_impact$revealed,
  " DEGs across key contrasts (B6J references).\n\n"
))
```

---

# What Genes Change With Treatment? {.tabset .tabset-pills}

**B6J references only.**

```{r volcano-setup}
main_effects <- c("HU_vs_NL_MAIN_EFFECT", "IR_vs_Sham_MAIN_EFFECT",
                  "KMP_vs_Sham_MAIN_EFFECT")
interactions <- c("Interaction_KMP_x_HU", "Interaction_KMP_x_IR",
                  "Interaction_HU_x_IR")
pairwise_contrasts <- c("HU_vs_Ctrl", "IR_vs_Ctrl", "KMP_vs_Ctrl")
volcano_contrasts <- c(main_effects, interactions, pairwise_contrasts)

# Color palette for significance change (main effects side-by-side)
change_pal <- c("Gained >1 mag." = "#2166AC",
                "Lost >1 mag." = "#B2182B",
                "Stable" = "#999999")
```

## Volcanos {.tabset .tabset-pills}

```{r volcano-tabs, results='asis', fig.width=12, fig.height=6}
for (ref in b6j_refs) {
  cat("\n\n###", ref, "{.tabset .tabset-pills}\n\n")

  res_adj_ref <- results_adj[[ref]]
  res_unadj_ref <- results_unadj[[ref]]
  if (is.null(res_adj_ref)) {
    cat("No adjusted results available.\n\n")
    next
  }

  for (contrast in volcano_contrasts) {
    clabel <- if (contrast %in% names(contrast_labels)) {
      contrast_labels[contrast]
    } else {
      contrast
    }
    cat("\n\n####", clabel, "\n\n")

    adj_df <- res_adj_ref[[contrast]]
    if (is.null(adj_df)) {
      cat("No results for this contrast.\n\n")
      next
    }

    adj_df <- adj_df |> left_join(gene_map, by = "gene_id")

    if (contrast %in% main_effects && !is.null(res_unadj_ref[[contrast]])) {
      # --- Side-by-side adjusted vs unadjusted (main effects) ---
      unadj_df <- res_unadj_ref[[contrast]]

      merged <- adj_df |>
        select(gene_id, log2FoldChange, padj, gene_name) |>
        inner_join(
          unadj_df |> select(gene_id, log2FoldChange, padj),
          by = "gene_id",
          suffix = c(".adj", ".unadj")
        ) |>
        mutate(
          neg_log10_adj = -log10(pmax(padj.adj, 1e-300)),
          neg_log10_unadj = -log10(pmax(padj.unadj, 1e-300)),
          delta = neg_log10_adj - neg_log10_unadj,
          change_cat = factor(case_when(
            delta > 1  ~ "Gained >1 mag.",
            delta < -1 ~ "Lost >1 mag.",
            TRUE ~ "Stable"
          ), levels = c("Gained >1 mag.", "Lost >1 mag.", "Stable"))
        )

      top_adj <- merged |>
        filter(!is.na(padj.adj) & padj.adj <= alpha) |>
        slice_min(padj.adj, n = 10, with_ties = FALSE)
      top_unadj <- merged |>
        filter(!is.na(padj.unadj) & padj.unadj <= alpha) |>
        slice_min(padj.unadj, n = 10, with_ties = FALSE)

      p_adj <- ggplot(merged, aes(x = log2FoldChange.adj, y = neg_log10_adj,
                                    color = change_cat)) +
        geom_point_rast(size = 0.8, alpha = 0.5, dpi = 600) +
        geom_hline(yintercept = -log10(alpha), linetype = "dashed", color = "grey40") +
        geom_vline(xintercept = 0, linetype = "dashed", color = "grey40") +
        geom_text_repel(
          data = top_adj, aes(label = gene_name),
          size = 3, segment.linetype = 2, max.overlaps = 15,
          color = "black", fontface = "italic"
        ) +
        scale_color_manual(values = change_pal) +
        labs(x = "log2 Fold Change", y = "-log10(padj)",
             subtitle = "Adjusted") +
        theme_report(base_size = 12) +
        theme(legend.position = "none")

      p_unadj <- ggplot(merged, aes(x = log2FoldChange.unadj, y = neg_log10_unadj,
                                      color = change_cat)) +
        geom_point_rast(size = 0.8, alpha = 0.5, dpi = 600) +
        geom_hline(yintercept = -log10(alpha), linetype = "dashed", color = "grey40") +
        geom_vline(xintercept = 0, linetype = "dashed", color = "grey40") +
        geom_text_repel(
          data = top_unadj, aes(label = gene_name),
          size = 3, segment.linetype = 2, max.overlaps = 15,
          color = "black", fontface = "italic"
        ) +
        scale_color_manual(values = change_pal) +
        labs(x = "log2 Fold Change", y = "-log10(padj)",
             subtitle = "Unadjusted") +
        theme_report(base_size = 12)

      p_combined <- p_adj | p_unadj
      print(p_combined + plot_layout(guides = "collect"))
      cat("\n\n")

    } else {
      # --- Single adjusted volcano (interactions + pairwise) ---
      volcano_df <- adj_df |>
        mutate(
          neg_log10_p = -log10(pmax(padj, 1e-300)),
          sig = ifelse(!is.na(padj) & padj <= alpha, "Sig", "NS")
        )

      top_genes <- volcano_df |>
        filter(sig == "Sig") |>
        slice_min(padj, n = 10, with_ties = FALSE)

      p_volc <- ggplot(volcano_df, aes(x = log2FoldChange, y = neg_log10_p,
                                        color = sig)) +
        geom_point_rast(size = 0.8, alpha = 0.5, dpi = 600) +
        geom_hline(yintercept = -log10(alpha), linetype = "dashed", color = "grey40") +
        geom_vline(xintercept = 0, linetype = "dashed", color = "grey40") +
        geom_text_repel(
          data = top_genes, aes(label = gene_name),
          size = 3, segment.linetype = 2, max.overlaps = 15,
          color = "black", fontface = "italic"
        ) +
        scale_color_manual(values = c("NS" = "#999999", "Sig" = "#ed9209")) +
        labs(x = "log2 Fold Change", y = "-log10(padj)", color = NULL) +
        theme_report(base_size = 12)
      print(p_volc)
      cat("\n\n")
    }
  }
}
```

## UpSet {.tabset .tabset-pills}

```{r upset-tabs, results='asis', fig.width=10, fig.height=6}
for (ref in b6j_refs) {
  cat("\n\n###", ref, "\n\n")

  res_adj_ref <- results_adj[[ref]]
  if (is.null(res_adj_ref)) {
    cat("No adjusted results available.\n\n")
    next
  }

  available_key <- intersect(key_contrasts, names(res_adj_ref))
  if (length(available_key) < 2) {
    cat("Fewer than 2 key contrasts available; skipping UpSet.\n\n")
    next
  }

  for (direction in c("up", "down")) {
    upset_list <- lapply(available_key, function(cname) {
      df <- res_adj_ref[[cname]]
      if (is.null(df)) return(character(0))
      if (direction == "up") {
        df$gene_id[!is.na(df$padj) & df$padj <= alpha & df$log2FoldChange > 0]
      } else {
        df$gene_id[!is.na(df$padj) & df$padj <= alpha & df$log2FoldChange < 0]
      }
    })
    names(upset_list) <- contrast_labels[available_key]

    all_genes <- unique(unlist(upset_list))
    if (length(all_genes) < 2) {
      cat(paste0("Fewer than 2 ", direction, "-regulated genes; skipping UpSet.\n\n"))
      next
    }

    upset_df <- data.frame(gene_id = all_genes)
    for (cname in names(upset_list)) {
      upset_df[[cname]] <- upset_df$gene_id %in% upset_list[[cname]]
    }

    bar_color <- if (direction == "up") "lightblue" else "#d15469"
    dir_label <- if (direction == "up") "Up-regulated" else "Down-regulated"

    p_upset <- upset(
      upset_df,
      names(upset_list),
      name = "Contrast",
      min_size = 1,
      min_degree = 2,
      base_annotations = list(
        "Intersection size" = intersection_size(
          bar_number_threshold = 1,
          fill = bar_color
        )
      ),
      themes = upset_default_themes(text = element_text(size = 8))
    ) + ggtitle(paste0(dir_label, " DEGs (adjusted) -- ", ref))
    print(p_upset)
    cat("\n\n")
  }
}
```

## Full Sig Gene Counts

```{r full-sig-table, results='asis'}
for (ref in b6j_refs) {
  cat("\n\n###", ref, "{.unlisted .unnumbered}\n\n")

  comp_ref <- comparisons |> filter(reference == ref)
  full_summary <- comp_ref |>
    group_by(contrast) |>
    summarise(
      `Sig. (adjusted)` = sum(padj.adj <= alpha, na.rm = TRUE),
      `Sig. (unadjusted)` = sum(padj.unadj <= alpha, na.rm = TRUE),
      `Adjusted only` = sum(adj_only_sig, na.rm = TRUE),
      `Unadjusted only` = sum(unadj_only_sig, na.rm = TRUE),
      Both = sum(both_sig, na.rm = TRUE),
      .groups = "drop"
    )
  print(styled_kable(full_summary,
                     caption = paste0("All 19 contrasts -- ", ref)))
  cat("\n\n")
}
```

---

# What Pathways Are Enriched? {.tabset .tabset-pills}

**B6J references only. Comparing adjusted and unadjusted models.**

```{r gsea-setup}
# Map contrast_slug to readable labels
slug_to_label <- contrast_labels
names(slug_to_label) <- gsub("[^A-Za-z0-9_]", "_", names(contrast_labels))

# Join adjusted and unadjusted GSEA results
gsea_joined <- data.frame()
if (nrow(gsea_summaries) > 0) {
  gsea_adj_only <- gsea_summaries |>
    filter(model == "adjusted") |>
    select(ID, Description, NES, p.adjust, database, reference, contrast_slug, contrast_name)

  gsea_unadj_only <- gsea_summaries |>
    filter(model == "unadjusted") |>
    select(ID, NES, p.adjust, database, reference, contrast_slug)

  gsea_joined <- gsea_adj_only |>
    inner_join(gsea_unadj_only,
               by = c("ID", "contrast_slug", "reference", "database"),
               suffix = c(".adj", ".unadj")) |>
    mutate(
      qscore_adj = -log10(pmax(p.adjust.adj, 1e-300)),
      qscore_unadj = -log10(pmax(p.adjust.unadj, 1e-300)),
      delta_q = qscore_adj - qscore_unadj,
      contrast_label = ifelse(
        contrast_slug %in% names(slug_to_label),
        slug_to_label[contrast_slug],
        contrast_name
      )
    )
}

# Helper to build a styled GT table
build_gsea_gt <- function(df, title = NULL, subtitle = NULL) {
  tbl <- df |>
    select(contrast_label, Description, p.adjust.adj, qscore_adj,
           p.adjust.unadj, qscore_unadj, delta_q)

  gt(tbl, groupname_col = "contrast_label", rowname_col = "Description") |>
    fmt_scientific(columns = c(p.adjust.adj, p.adjust.unadj), decimals = 2) |>
    fmt_number(columns = c(qscore_adj, qscore_unadj, delta_q), decimals = 1) |>
    cols_merge(
      columns = c(p.adjust.adj, qscore_adj),
      pattern = "{1} ({2})"
    ) |>
    cols_merge(
      columns = c(p.adjust.unadj, qscore_unadj),
      pattern = "{1} ({2})"
    ) |>
    cols_label(
      p.adjust.adj = "p.adjust (Q-score)",
      p.adjust.unadj = "p.adjust (Q-score)",
      delta_q = html("&Delta; Q-score")
    ) |>
    tab_spanner(label = "Cell-type adj.", columns = p.adjust.adj) |>
    tab_spanner(label = "Unadjusted", columns = p.adjust.unadj) |>
    cols_align(align = "center", columns = c(p.adjust.adj, p.adjust.unadj, delta_q)) |>
    opt_row_striping() |>
    cols_width(Description ~ px(300)) |>
    tab_style(
      style = list(
        cell_fill(color = "grey"),
        cell_text(weight = "bold")
      ),
      locations = cells_row_groups()
    ) |>
    tab_header(title = title, subtitle = subtitle)
}
```

## GO-BP {.tabset .tabset-pills}

```{r gsea-go-tabs, results='asis'}
if (nrow(gsea_joined) > 0) {
  for (ref in b6j_refs) {
    cat("\n\n###", ref, "\n\n")

    go_ref <- gsea_joined |>
      filter(reference == ref, database == "GO_BP")

    if (nrow(go_ref) == 0) {
      cat("No GO-BP GSEA results for this reference.\n\n")
      next
    }

    # Table 1: Top 5 terms per contrast (by adjusted Q-score)
    top_go <- go_ref |>
      group_by(contrast_label) |>
      slice_max(qscore_adj, n = 5, with_ties = FALSE) |>
      ungroup() |>
      arrange(contrast_label, desc(qscore_adj))

    if (nrow(top_go) > 0) {
      cat("\n\n**Top pathways per contrast (adjusted model)**\n\n")
      gt_top <- build_gsea_gt(top_go,
                              title = "GO-BP: Top Pathways per Contrast",
                              subtitle = ref)
      print(as_raw_html(gt_top))
      cat("\n\n")
    }

    # Table 2: Most changed pathways (by delta Q-score)
    # Top 10 gained (became more significant after adjustment)
    gained <- go_ref |>
      arrange(desc(delta_q)) |>
      slice_head(n = 10) |>
      mutate(contrast_label = paste0(contrast_label, " [gained]"))

    # Top 10 lost (became less significant after adjustment)
    lost <- go_ref |>
      arrange(delta_q) |>
      slice_head(n = 10) |>
      mutate(contrast_label = paste0(contrast_label, " [lost]"))

    most_changed <- bind_rows(gained, lost)

    if (nrow(most_changed) > 0) {
      cat("\n\n**Most changed pathways (adjusted vs unadjusted)**\n\n")
      gt_changed <- build_gsea_gt(most_changed,
                                  title = "GO-BP: Most Changed Pathways",
                                  subtitle = ref)
      print(as_raw_html(gt_changed))
      cat("\n\n")
    }
  }
} else {
  cat("\n\nNo GSEA results available.\n\n")
}
```

## KEGG {.tabset .tabset-pills}

```{r gsea-kegg-tabs, results='asis'}
if (nrow(gsea_joined) > 0) {
  for (ref in b6j_refs) {
    cat("\n\n###", ref, "\n\n")

    kegg_ref <- gsea_joined |>
      filter(reference == ref, database == "KEGG")

    if (nrow(kegg_ref) == 0) {
      cat("No KEGG GSEA results for this reference.\n\n")
      next
    }

    # Table 1: Top 5 terms per contrast (by adjusted Q-score)
    top_kegg <- kegg_ref |>
      group_by(contrast_label) |>
      slice_max(qscore_adj, n = 5, with_ties = FALSE) |>
      ungroup() |>
      arrange(contrast_label, desc(qscore_adj))

    if (nrow(top_kegg) > 0) {
      cat("\n\n**Top pathways per contrast (adjusted model)**\n\n")
      gt_top <- build_gsea_gt(top_kegg,
                              title = "KEGG: Top Pathways per Contrast",
                              subtitle = ref)
      print(as_raw_html(gt_top))
      cat("\n\n")
    }

    # Table 2: Most changed pathways
    gained <- kegg_ref |>
      arrange(desc(delta_q)) |>
      slice_head(n = 10) |>
      mutate(contrast_label = paste0(contrast_label, " [gained]"))

    lost <- kegg_ref |>
      arrange(delta_q) |>
      slice_head(n = 10) |>
      mutate(contrast_label = paste0(contrast_label, " [lost]"))

    most_changed <- bind_rows(gained, lost)

    if (nrow(most_changed) > 0) {
      cat("\n\n**Most changed pathways (adjusted vs unadjusted)**\n\n")
      gt_changed <- build_gsea_gt(most_changed,
                                  title = "KEGG: Most Changed Pathways",
                                  subtitle = ref)
      print(as_raw_html(gt_changed))
      cat("\n\n")
    }
  }
} else {
  cat("\n\nNo GSEA results available.\n\n")
}
```

---

# Session Info

<details>
<summary>Click to expand</summary>

```{r session-info}
sessionInfo()
```

</details>
